% discussion_final.tex — §7 Discussion

\paragraph{Is the thinking tax obvious in hindsight?}
While the \emph{mechanism} (truncation) may seem unsurprising, three aspects are not.
First, the \emph{magnitude} far exceeds what a simple format-overhead model would predict.
A naive account---``thinking wastes some tokens on tags and preamble, reducing the effective budget''---would predict a tax proportional to the format overhead: ${\sim}$10--20 tokens for \texttt{<think>} tags and preamble, i.e., a ${\sim}$5--8\% effective budget reduction at $b{=}256$.
This would predict at most a ${\sim}$5\,pp accuracy gap.
The observed gap is \textbf{69.5\,pp}---over $13\times$ larger.
The discrepancy arises because truncation is not a marginal tax but a \emph{catastrophic failure mode}: 98.6\% of thinking responses at $b{=}256$ produce \emph{zero parseable answer}, not merely a slightly degraded one.
Our decomposition (Proposition~\ref{prop:acc-decomp}) captures this: the tax scales with the \emph{entire truncation rate} $\rho(b) = 1 - F_L(b)$, not with a format constant.
A length-normalized comparison (accuracy per token) similarly fails to explain the phenomenon: nothink@256 achieves 87.5\% at 140 avg tokens (0.63\%/token), while think@256 achieves 18.0\% at 255 tokens (0.07\%/token)---a $9\times$ efficiency gap that no normalization can close.
Second, the \emph{tax amplification with chain length} (\S\ref{sec:finding:model-size}, Proposition~\ref{prop:inverse-scaling}) is counterintuitive: within the Qwen family, models that produce longer chains become \emph{worse} at thinking under the same budget, a phenomenon our decomposition explains through stochastic dominance of chain-length distributions.
No format-overhead argument predicts this: the overhead is constant across model sizes, but the tax amplifies $2.1\times$ from 8B to 9B/27B.
Third, the \emph{natural-stop oracle} (\S\ref{sec:finding:natural-stop}) is a novel signal with 99.0\% PPV (Proposition~\ref{prop:oracle}), not predicted by the truncation narrative alone.

\paragraph{Why hasn't this been studied before?}
The thinking tax has gone unnoticed because most reasoning evaluations impose generous budgets (4096--32768 tokens)---well above the crossover point---where thinking mode indeed dominates.
In practice, however, production deployments routinely operate at 256--1024 token budgets due to latency, cost, and throughput constraints~\citep{snell2024scaling}.
Our contribution is to systematically characterize this under-studied regime and show that the penalty is not marginal but \emph{severe}---rendering thinking mode strictly inferior across all tested conditions below the crossover.
The absence of this analysis from the literature reflects a gap between benchmark evaluation practice (unlimited budgets) and deployment reality (constrained budgets).

\paragraph{The format tax vs.\ the information tax.}
A key insight from the 27B nothink results (Appendix~\ref{app:27b-nothink}) resolves a natural question: is the tax due to model weakness, or format overhead?
The answer is definitively format: \textbf{27B nothink@512 achieves 95.5\%}---the \emph{highest} accuracy of any configuration we tested---while 27B think@512 achieves only 18.3\%.
The same model, answering the same questions, goes from near-perfect to near-random depending solely on whether it generates a reasoning chain.
This confirms that the thinking tax is a \emph{format tax}: the cost of wrapping answers in verbose reasoning traces that cannot fit within practical budgets.

\paragraph{When \emph{should} one think?}
Our crossover analysis (\S\ref{sec:theory}) provides a precise answer: thinking mode requires ${\sim}$2048 tokens on GSM8K (4$\times$ the non-thinking saturation budget) to match non-thinking accuracy; on MATH-500, the crossover demands an even larger budget, reflecting the longer chains harder problems produce.
Our theory (Proposition~\ref{prop:crossover}) explains this: the crossover budget $b^*$ corresponds to the $(\bar{\alpha}_{\mathrm{nt}}/\alpha_c)$-quantile of the chain-length distribution---approximately the 97th percentile.
Below these budgets, non-thinking mode is strictly preferred.
\method{} resolves this tension: triage difficulty via non-thinking probe,
allocate a split budget for reasoning, and decouple the answer from reasoning via multi-round self-distillation.

\paragraph{Why not simply use nothink with a larger budget?}
This is the right question.
On GSM8K, nothink accuracy plateaus at ${\sim}$89\% (nothink@256 $\approx$ nothink@512), so additional nothink budget provides no benefit, and \method{}'s structured think-then-extract pipeline achieves +5.0\,pp---the \emph{only} method to improve beyond the nothink plateau.
On MATH-500, the think budget ablation (\S\ref{sec:experiments:bthink}) provides decisive evidence.
Full-scale evaluation ($n{=}500$, same H800 hardware as baselines) provides definitive evidence.
IRIS@2048 achieves \textbf{67.2\%} \ci{63.0}{71.2} and IRIS@4096 achieves \textbf{74.0\%} \ci{70.0}{77.7}---exceeding nothink@1024 (59.8\%) by +7.4\,pp and +14.2\,pp respectively, with CI lower bounds well above the nothink@1024 point estimate.
The cross-budget gain of +6.8\,pp from $B_{\text{think}}{=}2048$ to 4096 (McNemar $p{=}0.0004$) confirms that the split-budget mechanism scales with the thinking budget.
At $B_{\text{think}}{=}4096$, 44.7\% of escalated chains complete naturally (vs.\ 8.8\% at 2048), and these achieve 71.7\% accuracy---validating that the model \emph{can} reason correctly when given sufficient budget.
The first 200 samples of each full-scale run reproduce the pilot exactly (73.0\% at 2048, 78.5\% at 4096), confirming methodological consistency across scale.

\paragraph{Implications for scaling and the 27B limitation.}
The tax amplification (Proposition~\ref{prop:inverse-scaling}) has troubling implications for models that produce longer chains.
Non-thinking accuracy is remarkably stable across model sizes---93.1\% (8B), 93.2\% (9B), 95.5\% (27B)---while thinking-mode accuracy \emph{collapses}: 56.9\%, 15.5\%, 18.3\% at the same budget.
The resulting tax grows from 36.2\,pp (8B) to 77.7\,pp (9B), a ${\sim}2.1\times$ amplification.
As models grow and chains lengthen (Corollary~\ref{cor:crossover-scaling}), the crossover budget $b^*$ will shift further right, making the thinking tax \emph{worse}, not better.

\textbf{Importantly, the cascade does not uniformly benefit all model sizes.}
At 27B scale with $B_1{=}256$, $B_2{=}512$, \town{} achieves only 62.8\%---\emph{below} \texttt{nothink@256} (65.1\%)---because the thinking fallback is too weak ($\text{think@512} = 18.3\%$ on routed samples).
The net recovery condition (Eq.~\ref{eq:net-recovery}) fails: more routing regrets than recoveries.
This provides a concrete \textbf{deployment criterion}: the cascade should only be applied when the fallback budget $B_2$ is large enough for thinking mode to add value on the routed hard subset.
For 27B, this requires $B_2 \geq 1024$ (where think accuracy rises to ${\sim}49\%$), implying a higher per-query cost for the fallback stage.
This scale-dependent asymmetry is a genuine limitation, and suggests that adaptive routing between thinking and non-thinking modes---rather than unconditional thinking---will become increasingly important for frontier models, but the \emph{minimum viable fallback budget} must be calibrated per model size.

\paragraph{Addressing potential concerns.}
We anticipate three objections and address them directly.

\emph{``Think vs.\ nothink compares two different policies, not just budget coupling.''}
True---and we address this with a matched-policy ablation.
IRIS and TOWN both use the \emph{same} thinking policy on the \emph{same} escalated samples; the only difference is whether reasoning and answering share a single budget (TOWN: coupled) or receive separate budgets (IRIS: decoupled).
On escalated MATH-500 samples, IRIS recovers +25.4\,pp over TOWN (Table~\ref{tab:escalated-ablation}, McNemar $p < 10^{-6}$).
This isolates the coupling mechanism: holding the thinking policy fixed, separating the answer budget recovers accuracy lost to truncation.
The think-vs-nothink comparison identifies the \emph{phenomenon}; the IRIS-vs-TOWN comparison isolates the \emph{cause}.

\emph{``The tax is just truncation/verbosity, not a reasoning issue.''}
Precisely---that is our point.
The \couplingtax{} is an \emph{architectural} inefficiency, not a reasoning failure.
When chains complete, accuracy is ${\sim}98\%$.
The problem is that coupling reasoning and answering under a shared budget means most chains \emph{never} complete.
Our truncation-waste decomposition (Proposition~\ref{prop:acc-decomp}) formalizes this: the tax is determined entirely by the chain-length CDF $F_L(b)$ and the conditional accuracies, not by reasoning quality.

\emph{``Does \method{} beat nothink@1024 on MATH-500?''}
Yes.
Full-scale evaluation ($n{=}500$, same H800 hardware as nothink@1024) provides a definitive answer: IRIS@2048 achieves \textbf{67.2\%} \ci{63.0}{71.2} (+7.4\,pp over nothink@1024's 59.8\%) and IRIS@4096 achieves \textbf{74.0\%} \ci{70.0}{77.7} (+14.2\,pp).
Both CI lower bounds exceed the nothink@1024 point estimate.
The cross-budget gain of +6.8\,pp from $B_{\text{think}}{=}2048$ to 4096 (McNemar $p{=}0.0004$) confirms monotonic scaling.
On GSM8K, where nothink has plateaued, \method{} is the only method that improves further (+5.0\,pp).

\emph{``The model scaling claim may be confounded by cross-model differences.''}
We control for this by using models within the same Qwen family (8B, 9B, 27B), sharing architecture, training pipeline, and \texttt{<think>} mode protocol.
The key evidence is that non-thinking accuracy is \emph{stable} across scales (93.1\%--95.5\%), while thinking accuracy collapses (56.9\% $\to$ 15.5\% $\to$ 18.3\%).
Since the same model achieves high accuracy in non-thinking mode, the collapse is attributable to longer chains, not weaker reasoning.

\paragraph{Production impact: the coupling tax as compute waste.}
The coupling tax is not merely an accuracy problem---it is a
\emph{compute waste} problem with direct production implications.
At $b{=}512$ on GSM8K, 62.0\% of thinking-mode responses are truncated,
consuming an average of 460 tokens per query while producing no
parseable answer.
For a service processing 1M queries/day at this budget, switching
from \texttt{think@512} to \texttt{nothink@512} eliminates
${\sim}$325M wasted tokens/day while \emph{increasing} accuracy by
36.2\,pp---a simultaneous improvement in both cost and quality.
On MATH-500, where truncation is even more severe (91.2\% at $b{=}2048$),
the waste scales proportionally.
Split-budget generation (IRIS) recovers value from these wasted
reasoning traces: rather than discarding a truncated chain that consumed
2048 tokens, the decoupled answer pass extracts a correct answer
from the partial reasoning in ${\sim}$200 additional tokens---turning
truncation waste into partial-reasoning credit.

\paragraph{Generalization and necessary conditions.}
Our experiments span three benchmark families: GSM8K, MATH-500, and five BIG-Bench Hard subtasks ($n{=}3{,}006$ total), three Qwen-family model sizes (8B, 9B, 27B), with cross-architecture validation on DeepSeek-R1-Distill-Llama-8B.
The tax appears on both easy and hard benchmarks---and is \emph{larger} on the harder one---while the crossover budget shifts predictably with task difficulty.
On BIG-Bench Hard (Appendix~\ref{app:bbh}), full-scale evaluation ($n{=}1{,}187$) confirms the tax persists in non-mathematical domains (logical deduction, object tracking, date understanding), with a +33.3\,pp gap at budget 256 and crossover between 1024--2048.

The coupling tax requires two necessary conditions:
(i)~\emph{truncation waste}: reasoning chains typically exceed the budget, causing most responses to produce no parseable answer; and
(ii)~\emph{a distinct alternative mode}: the model possesses an architecturally different non-thinking mode that can answer directly without generating a reasoning chain.
Condition~(i) is structural and model-agnostic---DeepSeek-R1-Distill-Llama-8B exhibits the same truncation patterns as Qwen models (Appendix~\ref{app:deepseek}).
Condition~(ii) is model-dependent: Qwen3's native \texttt{enable\_thinking=False} produces genuinely different (shorter, answer-focused) outputs, while DeepSeek's prompt-based nothink does not meaningfully alter generation behavior, yielding near-zero tax (${\leq}1.4$\,pp across all budgets).
This two-condition framing explains \emph{when} the tax manifests and \emph{why} its magnitude is architecture-dependent.

We explicitly scope our claims to \textbf{models with architecturally distinct thinking and non-thinking modes under fixed output-token budgets on structured reasoning tasks}.
The thinking tax may not apply (or may differ significantly) in three settings: (1)~tasks where CoT provides qualitatively different value, such as multi-hop retrieval, iterative code generation, or agentic workflows; (2)~closed-source models (e.g., OpenAI o1) where the thinking budget is managed internally; and (3)~settings with dynamic or unbounded token budgets where the crossover is always exceeded.
\method{}'s split-budget design demonstrates that decoupling reasoning from answer generation can mitigate the \couplingtax{}---confirmed on GSM8K (+5.0\,pp) and on MATH-500 at full scale ($n{=}500$): IRIS@4096 achieves 74.0\% \ci{70.0}{77.7}, exceeding nothink@1024 (59.8\%) by +14.2\,pp on identical hardware.
Extending split-budget generation with learned budget ratios, task-adaptive routing, and integration with multi-model ensembles is a promising direction for future work.
