% ===========================================================================
%  §5  Split-Budget Generation via MRSD
%  method_final.tex — NeurIPS 2026
%  Loaded via \input{sections/method_final}
% ===========================================================================

The decomposition in \S\ref{sec:theory} reveals a structural cause for the
thinking tax: reasoning and answering \emph{share a single output channel},
so that lengthy chains leave insufficient budget for the answer.
This section develops \emph{split-budget generation}, a principled
decoupling strategy, and instantiates it as
\textsc{Mrsd} (\textbf{M}ulti-\textbf{R}ound
\textbf{S}elf-\textbf{D}istillation), a training-free inference
framework that adaptively allocates compute across difficulty levels.
We present \method{} as a \textbf{principled mitigation} grounded in the
coupling-tax theory: since Proposition~\ref{prop:recoverable-tax} shows
that accuracy can be recovered by decoupling reasoning from answering,
\method{} instantiates this decoupling at inference time with zero
training overhead.
The deliberate simplicity is a feature: it demonstrates that the
coupling tax is recoverable through inference-time restructuring alone,
establishing a strong \emph{lower bound} on what more complex approaches
(learned routing, distillation, budget-aware fine-tuning) can achieve.

% ---------------------------------------------------------------------------
\subsection{The Coupling Problem}
\label{sec:method:coupling}
% ---------------------------------------------------------------------------

In standard thinking mode, reasoning tokens and answer tokens are
produced in a single autoregressive pass under one shared budget~$b$.
Let $Z = (z_1, \ldots, z_{|Z|})$ denote the reasoning trace and
$A = (a_1, \ldots, a_{|A|})$ the final answer; the hard constraint is:
\begin{equation}
  \label{eq:coupling-constraint}
  |Z| + |A| \;\le\; b.
\end{equation}
When the model's natural chain length $L(q) = |Z| + |A|$ exceeds~$b$,
generation is truncated, typically mid-chain before any answer tokens
are produced.
This creates a \emph{zero-sum coupling}: every reasoning token consumed
is one fewer token available for the answer.

Recall from Proposition~\ref{prop:acc-decomp} that thinking-mode accuracy
decomposes as:
\begin{equation}
  \mathrm{Acc}_{\mathrm{think}}(b)
  \;=\; \underbrace{\alpha_c \cdot F_L(b)}_{\text{completed chains}}
  \;+\; \underbrace{\alpha_t(b) \cdot \bigl(1 - F_L(b)\bigr)}_{\text{truncated chains}},
  \tag{\ref{eq:acc-decomp}}
\end{equation}
where $F_L(b) = \Pr(L \le b)$ is the chain-length CDF, $\alpha_c$ is the
accuracy of completed chains, and $\alpha_t(b) \ll \alpha_c$ is the
residual accuracy of truncated chains.
At budget 256 on GSM8K, $F_L(256) < 0.02$---over 98\% of chains are
truncated---making the first term negligible and the second term dominant
but small ($\alpha_t \approx 0$).

The coupling constraint in Eq.~\eqref{eq:coupling-constraint} is the
\emph{sole} mechanism behind the thinking tax: if reasoning and answering
could draw from separate budgets, truncated reasoning would not destroy
the model's ability to emit a correct answer.
This observation motivates the split-budget framework.

% ---------------------------------------------------------------------------
\subsection{Split-Budget Generation}
\label{sec:method:split}
% ---------------------------------------------------------------------------

\paragraph{Key insight.}
Decouple reasoning from answering by allocating \emph{separate} token
budgets to each stage.
Rather than requiring $|Z| + |A| \le b$ in a single pass, we introduce:
\begin{itemize}[nosep,leftmargin=*]
  \item A \textbf{reasoning pass}: generate a thinking trace with budget $B_r$,
    potentially allowing truncation;
  \item An \textbf{answer pass}: feed the (possibly truncated) reasoning
    trace as context to a non-thinking generation with budget $B_a$.
\end{itemize}
The total cost is $B_r + B_a$, but crucially, truncation of the reasoning
trace no longer prevents the model from emitting a well-formed answer.

\paragraph{Why truncated reasoning is still useful.}
A truncated chain $Z_{1:B_r}$ (with $B_r < L(q)$) is not worthless---it
contains partial problem decomposition, intermediate computations, and
subgoal completions.
Empirically, $\alpha_t(b)$ increases from ${\approx}0$ at $b{=}128$ to 31.8\%
at $b{=}512$ (Table~\ref{tab:natural-stop-oracle}), demonstrating that
even incomplete reasoning carries substantial information about the answer.
By feeding $Z_{1:B_r}$ as a \emph{context prefix} to a non-thinking pass
that generates the answer independently, the model can:
(i)~read and integrate the partial reasoning,
(ii)~fill in any missing steps, and
(iii)~produce a complete, well-formed answer within $B_a$ tokens.

\paragraph{Formal decomposition.}
Let $\alpha_{\mathrm{split}}(B_r, B_a)$ denote the accuracy of
split-budget generation.
This replaces the coupled accuracy with:
\begin{equation}
  \label{eq:split-acc}
  \alpha_{\mathrm{split}}(B_r, B_a)
  \;=\; F_L(B_r) \cdot \alpha_c
  \;+\; \bigl(1 - F_L(B_r)\bigr) \cdot
    \alpha_{\mathrm{extract}}(B_r, B_a),
\end{equation}
where $\alpha_{\mathrm{extract}}(B_r, B_a) \gg \alpha_t(B_r)$ is the
accuracy of the decoupled answer pass on truncated chains.
The gain over coupled thinking at the same reasoning budget $B_r$
(Proposition~\ref{prop:recoverable-tax}) is:
\begin{equation}
  \label{eq:split-gain}
  \Delta \;=\; \bigl(1 - F_L(B_r)\bigr) \cdot
    \bigl(\alpha_{\mathrm{extract}}(B_r, B_a) - \alpha_t(B_r)\bigr)
  \;\ge\; 0,
\end{equation}
which is positive whenever the decoupled answer pass improves upon the
truncation residual---a mild condition satisfied in all our experiments.
The gain is largest when $F_L(B_r) \ll 1$ (heavy truncation) and
$\alpha_{\mathrm{extract}} \gg \alpha_t$ (large recovery margin),
precisely the regime where the coupling tax is most severe.

% ---------------------------------------------------------------------------
\subsection{\textsc{Mrsd}: Multi-Round Self-Distillation}
\label{sec:method:mrsd}
% ---------------------------------------------------------------------------

Split-budget generation recovers value from truncated chains in a single
round.
For problems where one round of reasoning is insufficient, we extend the
framework to \textsc{Mrsd}: an iterative procedure that progressively
refines the answer through multiple think-then-extract rounds, with
each round informed by the previous attempt.
The full procedure is given in Algorithm~\ref{alg:mrsd}.

\begin{algorithm}[t]
\caption{\textsc{Mrsd}: Multi-Round Self-Distillation}
\label{alg:mrsd}
\begin{algorithmic}[1]
\Require Query $q$, model $\mathcal{M}$,
  triage budget $B_1$, thinking budget $B_r$,
  answer budget $B_a$, max rounds $K$
\Ensure Answer $\hat{a}$, total tokens used $T$

\medskip
\Statex \textbf{Stage 0: Difficulty triage}
\State $y_0 \leftarrow \mathcal{M}_{\mathrm{nothink}}(q,\; B_1)$
  \Comment{Non-thinking probe}
\If{$|y_0| < B_1$}
  \Comment{Natural stop $\Rightarrow$ easy; accept immediately}
  \State \Return $\mathrm{answer}(y_0),\; |y_0|$
\EndIf

\medskip
\Statex \textbf{Stage 1: First think-then-extract round}
\State $Z_1 \leftarrow \mathcal{M}_{\mathrm{think}}(q,\; B_r)$
  \Comment{Generate reasoning trace (may truncate)}
\State $y_1 \leftarrow \mathcal{M}_{\mathrm{nothink}}(q \,\|\, Z_1,\; B_a)$
  \Comment{Decoupled answer extraction}
\State $\hat{a}_1 \leftarrow \mathrm{answer}(y_1)$

\medskip
\Statex \textbf{Stages $k = 2, \ldots, K$: Iterative refinement}
\For{$k = 2, \ldots, K$}
  \State $\mathrm{hint}_k \leftarrow \hat{a}_{k-1}$
    \Comment{Previous answer as hint}
  \State $Z_k \leftarrow \mathcal{M}_{\mathrm{think}}(
    q \,\|\, \text{``Previous attempt: } \mathrm{hint}_k \text{''},\; B_r)$
  \State $y_k \leftarrow \mathcal{M}_{\mathrm{nothink}}(q \,\|\, Z_k,\; B_a)$
  \State $\hat{a}_k \leftarrow \mathrm{answer}(y_k)$
  \If{$\hat{a}_k = \hat{a}_{k-1}$}
    \Comment{Convergence: consecutive agreement}
    \State \Return $\hat{a}_k, \; B_1 + k \cdot (B_r + B_a)$
  \EndIf
\EndFor

\medskip
\Statex \textbf{Fallback: majority vote}
\State \Return $\mathrm{majority}(\hat{a}_1, \ldots, \hat{a}_K),
  \; B_1 + K \cdot (B_r + B_a)$
\end{algorithmic}
\end{algorithm}

\paragraph{Stage 0: Difficulty triage via non-thinking probe.}
We first attempt the query in non-thinking mode with a modest triage
budget $B_1$ (default: 256 tokens).
If the model produces a complete answer before exhausting $B_1$---a
\emph{natural stop}---we accept it immediately.
This exploits Finding~1's observation that non-thinking mode dominates
at low budgets: on GSM8K, ${\sim}$88.8\% of queries terminate naturally
at $B_1{=}256$, with an average cost of only 133 tokens and accuracy
matching \texttt{nothink@256} (87.5\%).
The natural-stop signal functions as a high-precision difficulty filter
(Proposition~\ref{prop:oracle}): questions that complete are
overwhelmingly easy, making escalation unnecessary.

\paragraph{Stage 1: Think-then-extract.}
For the ${\sim}$11.2\% of queries that exhaust the triage budget (hard
problems), we invoke split-budget generation:
\begin{enumerate}[nosep,leftmargin=*]
  \item \textbf{Think@$B_r$}: Generate a reasoning trace in thinking
    mode with budget $B_r$.
    The trace may be truncated, but this is acceptable---truncated
    reasoning still carries useful intermediate information.
  \item \textbf{Nothink@$B_a$}: Concatenate the query~$q$ and reasoning
    trace~$Z_1$ as context, then generate the answer in non-thinking
    mode with a dedicated budget $B_a$.
    The answer is produced by a clean, non-thinking pass that reads
    the reasoning as a ``scratchpad'' but is never constrained by its length.
\end{enumerate}
This decoupling is the core innovation: reasoning tokens and answer
tokens draw from \emph{separate budgets}, eliminating the zero-sum
coupling that causes the thinking tax.

\paragraph{Stages $k \ge 2$: Iterative refinement with hints.}
When the first round fails to produce a correct answer (or when
additional confidence is desired), subsequent rounds provide the
previous answer as a \emph{hint} to the thinking pass:
\begin{quote}
  \texttt{Previous attempt: $\hat{a}_{k-1}$. Re-examine the problem
  and verify or correct this answer.}
\end{quote}
The thinking model can then:
(i)~verify the previous answer and reinforce it if correct,
(ii)~identify errors and explore alternative solution paths, or
(iii)~decompose the problem differently given the hint.
The resulting trace $Z_k$ is again fed to a non-thinking answer
extraction pass, producing $\hat{a}_k$.

\paragraph{Convergence criterion.}
We stop iterating when two consecutive rounds produce the same
answer: $\hat{a}_k = \hat{a}_{k-1}$.
Agreement signals that the model has converged to a stable solution.
Empirically, 98\% of escalated queries converge within $K{=}3$ rounds
on GSM8K (pilot, $n{=}200$).

\paragraph{Fallback: majority vote.}
If the maximum number of rounds $K$ is reached without convergence,
we take a majority vote over all round answers
$\{\hat{a}_1, \ldots, \hat{a}_K\}$.
This provides a robust aggregation that benefits from the diversity of
reasoning traces across rounds.

% ---------------------------------------------------------------------------
\subsection{Analysis of Token Efficiency}
\label{sec:method:cost}
% ---------------------------------------------------------------------------

We analyze the expected token cost of \textsc{Mrsd} and compare it
against fixed-mode baselines and self-consistency.

\paragraph{Expected cost decomposition.}
Let $p$ denote the fraction of queries resolved at Stage~0
(natural-stop rate at budget $B_1$), and let $\bar{K}$ denote the
average number of refinement rounds for escalated queries.
The expected per-query token cost is:
\begin{equation}
  \label{eq:mrsd-cost}
  \mathbb{E}[T_{\textsc{Mrsd}}]
  \;=\; p \cdot \bar{t}_1
  \;+\; (1-p) \cdot \bigl(B_1 + \bar{K} \cdot (B_r + B_a)\bigr),
\end{equation}
where $\bar{t}_1 \le B_1$ is the average token cost of queries that
stop naturally.
On GSM8K with $B_1{=}256$, $B_r{=}512$, $B_a{=}128$: $p = 0.888$,
$\bar{t}_1 = 133$, $\bar{K} \approx 1.5$, giving
$\mathbb{E}[T] = 0.888 \times 133 + 0.112 \times (256 + 1.5 \times 640)
\approx 118 + 136 = 254$ tokens per query (observed: 235, as early
convergence and short answers reduce the actual token count below
worst-case budgets).

\paragraph{Comparison with baselines.}
We compare cost structures across methods:
\begin{itemize}[nosep,leftmargin=*]
  \item \textbf{\texttt{nothink@$b$}}: Fixed non-thinking budget.
    Cost $= b$ per query (no adaptation).
    Accuracy saturates at moderate budgets but cannot leverage reasoning
    for hard problems.
  \item \textbf{\texttt{think@$b$}}: Fixed thinking budget.
    Cost $= b$ per query.
    Accuracy limited by truncation waste when $b < b^*$
    (Proposition~\ref{prop:crossover}).
  \item \textbf{Self-consistency (SC@$k$)}: Sample $k$ independent
    non-thinking responses and take a majority vote.
    Cost $= k \cdot b$ per query.
    Improves accuracy through diversity but scales linearly in~$k$
    with diminishing returns.
  \item \textbf{\town{} cascade}: Two-stage triage (nothink@$B_1$ then
    think@$B_2$).
    Cost $= p \bar{t}_1 + (1-p)(B_1 + B_2)$.
    Efficient triage, but the escalation stage still suffers from the
    coupling problem.
  \item \textbf{\textsc{Mrsd}}: Adaptive triage + decoupled reasoning.
    Cost given by Eq.~\eqref{eq:mrsd-cost}.
    Spends tokens efficiently by (i)~resolving easy queries cheaply,
    (ii)~decoupling reasoning from answering, and (iii)~iterating only
    when needed.
\end{itemize}

\paragraph{Why \textsc{Mrsd} is more token-efficient.}
The key advantage is \emph{adaptive allocation}: easy queries ($p \approx 89\%$)
cost only $\bar{t}_1 \approx 133$ tokens, while hard queries receive
multi-round reasoning up to $K(B_r + B_a)$ tokens.
Compared to SC@$k$ at matched total budget, \textsc{Mrsd} invests
its tokens in \emph{complementary} ways---combining reasoning-mode
depth with non-thinking answer extraction---rather than sampling
$k$ independent shallow responses.
Compared to \town{}, the decoupled answer pass converts truncated
reasoning (previously ${\sim}0\%$ accuracy at low budgets) into usable
signal, achieving Stage~3 accuracy of 76.9\% on escalated samples
vs.\ only 15.4\% for standard truncated thinking
(pilot, $n{=}200$).

\paragraph{Interpolation dominance.}
Both \town{} and \textsc{Mrsd} strictly improve on at least one
objective relative to each single-mode baseline:

\begin{theorem}[\textsc{Mrsd} interpolation dominance]
\label{thm:mrsd-pareto}
If the \emph{net recovery condition}
\begin{equation}
  \label{eq:net-recovery}
  \alpha_{\mathrm{extract}}^{\mathrm{hard}} > \alpha_{\mathrm{nt}}^{\mathrm{hard}}
\end{equation}
holds (split-budget generation on escalated queries achieves higher
accuracy than non-thinking alone) and the triage early-stop rate
$p > 0$, then \textsc{Mrsd} achieves:
\begin{enumerate}[nosep,leftmargin=*]
  \item Higher accuracy than \texttt{nothink@}$B_1$, and
  \item Lower expected cost than \texttt{think@}$(B_1 + K(B_r + B_a))$.
\end{enumerate}
On GSM8K (pilot, $n{=}200$), the condition is satisfied:
$\alpha_{\mathrm{extract}}^{\mathrm{hard}} = 87.5\%$ (14/16 escalated
queries rescued) vs.\ $\alpha_{\mathrm{nt}}^{\mathrm{hard}} \approx 0\%$.

\textbf{Caveat:} the accuracy dominance is over \texttt{nothink@}$B_1$---the triage budget.
If nothink accuracy improves substantially when $B_1$ is doubled
(as on MATH-500: 47.5\% $\to$ 59.8\% full-scale at $n{=}500$), it may be more efficient
to increase the nothink budget than to deploy split-budget generation.
\method{} is most effective when nothink accuracy has saturated at $B_1$.
\end{theorem}

\begin{proof}
For accuracy: decompose both methods by the triage event
$\mathcal{S} = \{|y_0| < B_1\}$.
Nothink accuracy: $\mathrm{Acc}_{\mathrm{nt}}(B_1) = p \cdot \alpha_c^{\mathrm{nt}}
+ (1-p) \cdot \alpha_{\mathrm{nt}}^{\mathrm{hard}}$.
\textsc{Mrsd} accuracy:
$\mathrm{Acc}_{\textsc{Mrsd}} = p \cdot \alpha_c^{\mathrm{nt}}
+ (1-p) \cdot \alpha_{\mathrm{extract}}^{\mathrm{hard}}$.
Subtracting: $\mathrm{Acc}_{\textsc{Mrsd}} - \mathrm{Acc}_{\mathrm{nt}}(B_1)
= (1-p)(\alpha_{\mathrm{extract}}^{\mathrm{hard}} - \alpha_{\mathrm{nt}}^{\mathrm{hard}}) > 0$
by the net recovery condition (full proof in Appendix~\ref{app:town-proof}).
For cost: the expected cost (Eq.~\ref{eq:mrsd-cost}) satisfies
$\mathbb{E}[T] = p \bar{t}_1 + (1-p)(B_1 + \bar{K}(B_r + B_a))
< B_1 + K(B_r + B_a)$ whenever $p > 0$, since $\bar{t}_1 < B_1$
for early-stop queries.
\end{proof}

% ---------------------------------------------------------------------------
\subsection{Design Choices and Practical Considerations}
\label{sec:method:design}
% ---------------------------------------------------------------------------

\paragraph{Budget allocation.}
The triage budget $B_1$ should be large enough to accommodate most
answer lengths (we use $B_1{=}256$, sufficient for ${\sim}$89\% of
GSM8K answers).
The reasoning budget $B_r$ controls the depth of thinking; even at
moderate values ($B_r{=}512$), truncated traces contain sufficient
intermediate information for the extraction pass.
The answer budget $B_a$ can be relatively small ($B_a{=}128$) because
the extraction pass leverages the reasoning context and only needs
to produce the final answer.

\paragraph{Hint construction.}
In refinement rounds ($k \ge 2$), we prepend a minimal hint to the
query: the previous numerical answer and an instruction to verify or
correct it.
We avoid feeding the full previous reasoning trace as a hint, as this
would consume budget without guaranteed benefit; the compact answer
serves as a sufficient ``seed'' for re-examination.

\paragraph{Convergence vs.\ fixed rounds.}
The convergence criterion (consecutive agreement) provides an adaptive
stopping rule that avoids unnecessary computation on problems that
converge quickly.
In the pilot study, 72\% of escalated queries converge after Round~1,
and 98\% by Round~3.
An alternative is to fix $K$ and always majority-vote; we find
convergence-based stopping reduces average cost by ${\sim}$20\% with
no accuracy loss.

\paragraph{Compatibility with existing infrastructure.}
\textsc{Mrsd} requires no model modification, fine-tuning, or access
to internal activations.
It uses only the model's standard
\texttt{enable\_thinking=True/False} API, making it applicable to any
reasoning model with dual-mode support (e.g., Qwen3, DeepSeek-R1).
The multi-round procedure can be implemented as a simple wrapper around
any inference API, with each round being an independent forward pass.
