\section{Overthinking: Definition and Prevalence}

\subsection{Problem Setup}

We study reasoning tasks where a language model $M$ generates a solution $y$ given a question $x$ and a token budget $b$.
Let $\text{Acc}(M, x, b)$ denote whether the model produces a correct answer under budget $b$.

\begin{definition}[Overthinking]
A question $x$ exhibits \textbf{overthinking} at budgets $(b_1, b_2)$ with $b_1 < b_2$ if:
$$\text{Acc}(M, x, b_1) = 1 \quad \text{and} \quad \text{Acc}(M, x, b_2) = 0$$
\end{definition}

Intuitively, overthinking occurs when a model solves a problem correctly with limited compute but fails when given more resources.

\subsection{Experimental Setup}

\paragraph{Models.} We evaluate two Qwen models: Qwen3-8B and Qwen3.5-27B~\cite{qwen2024}.

\paragraph{Benchmarks.}
\begin{itemize}[nosep]
\item \textbf{GSM8K}~\cite{cobbe2021gsm8k}: Grade-school math (1,319 problems)
\item \textbf{MATH500}~\cite{hendrycks2021math}: Competition math (500 problems)
\item \textbf{BIG-Bench Hard}~\cite{suzgun2022bbh}: Multi-domain reasoning (6,511 problems)
\end{itemize}

\paragraph{Budgets.} We test three token budgets: $b \in \{128, 256, 512\}$ for GSM8K, and $\{512, 1024, 2048\}$ for MATH500.

\paragraph{Evaluation protocol.} For each $(model, benchmark, budget)$ combination, we run multiple random seeds and aggregate results.
GSM8K-27B uses 23 seeds (920 samples), GSM8K-8B uses 7 seeds (280 samples).

\subsection{Main Findings}

Table~\ref{tab:overthinking_rates} shows overthinking rates across settings.

\begin{table}[h]
\centering
\caption{Overthinking prevalence: percentage of samples correct at lower budget but wrong at higher budget.}
\label{tab:overthinking_rates}
\begin{tabular}{lccc}
\toprule
Setting & Correct@256 & Overthinking & Rate \\
\midrule
GSM8K-27B & 425/920 & 229/425 & 33.8\% \\
GSM8K-8B & 173/280 & 58/173 & 33.5\% \\
\midrule
MATH500-8B & 90/500 & 31/90 & 34.4\% \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Key observations:}
\begin{enumerate}[nosep]
\item Overthinking is \emph{common}: 30-40\% of correctly-solved problems degrade with more compute
\item The phenomenon is \emph{consistent} across model scales (8B vs 27B)
\item It generalizes across task types (math word problems vs competition math)
\end{enumerate}
