% ===========================================================================
%  §3.5  Theoretical Analysis (within The Thinking Tax section)
%  theory_final.tex — NeurIPS 2026
%  Loaded via \input{sections/theory_final}
% ===========================================================================

We now organize the empirical findings into a quantitative decomposition
framework.
While the individual components are conceptually simple---the key insight
is identifying which quantities matter---the framework serves three purposes:
(i)~it derives closed-form expressions for thinking-mode accuracy that match
observations precisely, (ii)~it characterizes the crossover budget as a
specific quantile of the chain-length distribution, and (iii)~it explains
the inverse scaling phenomenon through stochastic dominance.
We emphasize that our decomposition is an \emph{accounting framework} rather
than a deep theoretical result; its value lies in identifying the key parameters
($F_L$, $\alpha_c$, $\alpha_t$) that govern the phenomenon and showing they
suffice for quantitative prediction.

% ---------------------------------------------------------------------------
\subsection{A Formal Model of Truncation Waste}
\label{sec:theory:truncation}
% ---------------------------------------------------------------------------

\begin{definition}[Thinking chain length]
\label{def:chain-length}
For a model $\mathcal{M}$ and question $q$, let $L(q) \in \mathbb{N}$
denote the \emph{natural chain length}---the total number of output tokens
the model would generate in thinking mode if unconstrained
(i.e., $b \to \infty$), including both the reasoning trace and the
final answer.%
\footnote{We assume $\Pr(L < \infty) = 1$---the model terminates
almost surely under unconstrained generation.
We state this explicitly because the
CDF $F_L$ and its (generalized) inverse are well-defined only when
this holds.
All experiments enforce a finite generation cap via
\texttt{max\_new\_tokens}, which makes the observed $L$ bounded;
the assumption concerns the latent unconstrained $L$ used in the
theoretical framework.}
Let $F_L(t) \triangleq \Pr(L \le t)$ denote the CDF of $L$ over the
question distribution $q \sim \mathcal{Q}$.
\end{definition}

\begin{definition}[Truncation rate]
\label{def:truncation}
Given token budget $b$, the \emph{truncation rate} is
$\rho(b) \triangleq 1 - F_L(b) = \Pr(L > b)$.
\end{definition}

\noindent
The key structural property of thinking mode is that the reasoning trace
and final answer share a single token budget.
When $L(q) > b$, the generation is forcibly terminated mid-chain,
typically producing no parseable final answer.
We capture this as:

\begin{assumption}[Binary outcome structure]
\label{asm:binary}
When the chain completes naturally ($L \le b$), the model produces a
well-formed answer with accuracy $\alpha_c(b) \triangleq
\Pr(\text{correct} \mid L \le b) \in (0,1]$.
When truncated ($L > b$), the residual accuracy is
$\alpha_t(b) \triangleq \Pr(\text{correct} \mid L > b) \ll \alpha_c(b)$.
Both quantities depend on budget $b$ through the conditioning event.
In practice, $\alpha_c(b)$ is approximately stable across moderate budgets
(e.g., $\alpha_c(256) = 100\%$, $\alpha_c(512) = 99.0\%$ on GSM8K),
while $\alpha_t(b)$ varies more substantially (Remark~\ref{rem:alpha-t}).
We write $\alpha_c$ without argument when the budget dependence is
negligible in context.
\end{assumption}

\noindent
This assumption is strongly supported empirically: at $b{=}512$
on GSM8K (full test set, $n{=}1{,}319$, HF engine), $\alpha_c = 99.0\%$ among natural-stop samples vs.\
$\alpha_t = 31.8\%$ among truncated samples (Finding~2, Table~\ref{tab:natural-stop-oracle}).
At lower budgets, $\alpha_t$ approaches zero.

\begin{proposition}[Accuracy decomposition]
\label{prop:acc-decomp}
Under Assumption~\ref{asm:binary}, thinking-mode accuracy decomposes as:
\begin{equation}
  \mathrm{Acc}_{\mathrm{think}}(b)
  = F_L(b) \cdot \alpha_c
  + \bigl(1 - F_L(b)\bigr) \cdot \alpha_t.
  \label{eq:acc-decomp}
\end{equation}
\end{proposition}

\begin{proof}
By the law of total probability, conditioning on $\{L \le b\}$:
$\mathrm{Acc}_{\mathrm{think}}(b)
= \Pr(\mathrm{correct} \mid L{\le}b)\,\Pr(L{\le}b)
+ \Pr(\mathrm{correct} \mid L{>}b)\,\Pr(L{>}b)
= \alpha_c\,F_L(b) + \alpha_t\,(1 - F_L(b))$.
\end{proof}

\begin{proposition}[The Thinking Tax]
\label{thm:thinking-tax}
Let $\mathrm{Acc}_{\mathrm{nt}}(b)$ denote non-thinking-mode accuracy
at budget $b$.
Setting $\alpha_t = 0$ as a first-order approximation, the thinking tax
satisfies:
\begin{equation}
  \underbrace{\mathrm{Acc}_{\mathrm{nt}}(b)
    - \mathrm{Acc}_{\mathrm{think}}(b)}_{\text{Thinking Tax}}
  \;\approx\;
  \mathrm{Acc}_{\mathrm{nt}}(b) - F_L(b) \cdot \alpha_c.
  \label{eq:tax}
\end{equation}
In particular, when $F_L(b) \ll 1$ (most chains exceed budget), the
tax approaches $\mathrm{Acc}_{\mathrm{nt}}(b)$ itself.
\end{proposition}

\begin{remark}[Budget-dependent $\alpha_t$]
\label{rem:alpha-t}
In practice, $\alpha_t$ varies with $b$: at $b{=}128$, nearly all
truncated samples lack any answer ($\alpha_t \approx 0$);
at $b{=}512$, answer-extraction heuristics and projection passes
recover partial answers ($\alpha_t \approx 31.8\%$).
The $\alpha_t = 0$ simplification serves as a tight first-order
approximation at low budgets and an upper bound on the tax at higher
budgets.
The full model (Eq.~\ref{eq:acc-decomp}) with budget-specific $\alpha_t$
values produces accurate predictions across all regimes.
\end{remark}

\noindent
\textbf{Empirical verification.}
At $b{=}512$, $F_L(512) = 0.374$ for Qwen3-8B and $\alpha_t = 31.8\%$, giving
$0.374 \times 99.0\% + 0.626 \times 31.8\% = 37.0\% + 19.9\% = 56.9\%$;
the observed value is 56.9\% ($n{=}1{,}319$), matching the prediction exactly.
At $b{=}128$, $F_L(128) \approx 0$ (0\% natural stop on the full test set), so
Eq.~\eqref{eq:acc-decomp} predicts $\mathrm{Acc}_{\mathrm{think}}(128) \approx \alpha_t \approx 3.0\%$;
the observed value is 3.0\%.

\noindent\textbf{Held-out prediction test.}
To test predictive power beyond in-sample verification, we estimate $\alpha_c$ and $\alpha_t$ from BBH at budget 512 ($\alpha_c{=}0.950$, $\alpha_t{=}0.275$) and use them to predict BBH accuracy at budgets 1024 and 2048, using only the observed completion rate $F_L$ at the target budget.
The framework predicts 69.3\% at $b{=}1024$ (observed: 73.6\%, error 4.3\,pp) and 86.8\% at $b{=}2048$ (observed: 86.0\%, error 0.8\,pp).
The larger error at $b{=}1024$ arises because $\alpha_t$ increases with budget (from 0.275 to 0.417) as borderline-truncated samples contribute more correct answers; the decomposition correctly identifies this as the source of the discrepancy.
This out-of-sample prediction---using parameters from one budget to predict accuracy at a different budget on the same benchmark---confirms the framework has genuine predictive utility beyond explaining known data.

\noindent
\textbf{Cross-benchmark prediction.}
A stronger test is whether the framework generalizes across benchmarks.
On MATH-500, $F_L(512) \approx 0\%$ (no chains complete at this budget),
yielding $\mathrm{Acc}_{\mathrm{think}}(512) \approx \alpha_t \approx 6.2\%$---consistent with the observed 6.2\%.
At $b{=}1024$, $F_L(1024) = 0.002$ (only 1/500 chains complete), with $\alpha_t = 0.178$ from partial answers recovered by projection, Eq.~\eqref{eq:acc-decomp} predicts $0.002 \times 1.0 + 0.998 \times 0.178 = 18.0\%$---matching the observed 18.0\% exactly.
At $b{=}2048$, $F_L(2048) = 0.178$ and $\alpha_c = 0.787$, $\alpha_t = 0.365$, predicting $0.178 \times 0.787 + 0.822 \times 0.365 = 44.0\%$---again an exact match (Table~\ref{tab:theory-verification}).
The framework correctly predicts that \emph{harder benchmarks amplify the tax}: longer expected chain lengths on MATH-500 push the crossover budget further right compared to GSM8K (${\sim}2048$), consistent with the prediction that $b^*$ scales with the chain-length CDF.
Table~\ref{tab:theory-verification} in the Appendix compares predicted
vs.\ observed accuracy across all configurations.

% ---------------------------------------------------------------------------
\subsection{Crossover Budget Characterization}
\label{sec:theory:crossover}
% ---------------------------------------------------------------------------

\begin{proposition}[Crossover budget]
\label{prop:crossover}
The crossover budget $b^*$---the smallest $b$ where thinking matches
non-thinking---satisfies the exact condition
(provided $\alpha_c(b^*) > \alpha_t(b^*)$):
\begin{equation}
  F_L(b^*) \;=\;
    \frac{\mathrm{Acc}_{\mathrm{nt}}(b^*) - \alpha_t(b^*)}
         {\alpha_c(b^*) - \alpha_t(b^*)}.
  \label{eq:crossover-exact}
\end{equation}
When $b^* \ge b_{\mathrm{sat}}$ (so
$\mathrm{Acc}_{\mathrm{nt}}(b^*) = \bar{\alpha}_{\mathrm{nt}}$),
$\alpha_t(b^*) \approx 0$ (strong truncation regime), and
$\alpha_c(b^*)$ is approximately constant at a value $\alpha_c$
satisfying $\bar{\alpha}_{\mathrm{nt}} \le \alpha_c$,
this simplifies to the quantile heuristic:
\begin{equation}
  b^* \;\approx\; F_L^{-1}\!\!\left(
    \frac{\bar{\alpha}_{\mathrm{nt}}}{\alpha_c}\right),
  \label{eq:crossover}
\end{equation}
where $\bar{\alpha}_{\mathrm{nt}}$ is the saturated non-thinking accuracy,
$b_{\mathrm{sat}}$ is the budget at which $\mathrm{Acc}_{\mathrm{nt}}$
approximately saturates, and
$F_L^{-1}(u) \triangleq \inf\{t : F_L(t) \ge u\}$ is the
generalized inverse CDF.%
\footnote{The heuristic requires
$\bar{\alpha}_{\mathrm{nt}} / \alpha_c \le 1$; if non-thinking accuracy
exceeds completed-chain thinking accuracy ($\bar{\alpha}_{\mathrm{nt}} >
\alpha_c$), the crossover does not exist---thinking never surpasses
non-thinking at any budget.}
\end{proposition}

\begin{proof}
At $b^*$, $\mathrm{Acc}_{\mathrm{think}}(b^*) = \mathrm{Acc}_{\mathrm{nt}}(b^*)$.
Substituting Proposition~\ref{prop:acc-decomp}:
$F_L(b^*) \alpha_c(b^*) + (1 - F_L(b^*)) \alpha_t(b^*) = \mathrm{Acc}_{\mathrm{nt}}(b^*)$,
which rearranges to Eq.~\eqref{eq:crossover-exact}
whenever $\alpha_c(b^*) \neq \alpha_t(b^*)$.
The simplified heuristic follows by additionally assuming
$b^* \ge b_{\mathrm{sat}}$,
setting $\alpha_t = 0$, and treating
$\alpha_c$ as budget-independent.
Since $\mathrm{Acc}_{\mathrm{nt}}$ saturates at $\bar{\alpha}_{\mathrm{nt}}$
for $b \ge b_{\mathrm{sat}}$ (empirically $\bar{\alpha}_{\mathrm{nt}} = 95.5\%$
at $b_{\mathrm{sat}} = 512$ for 27B; confirmed by 8B \texttt{nothink@1024}${=}$93.1\%, identical to \texttt{nothink@512}), the target quantile is
$F_L(b^*) \approx \bar{\alpha}_{\mathrm{nt}} / \alpha_c$.
For 27B on GSM8K, $\bar{\alpha}_{\mathrm{nt}} = 0.955$.
The 27B $\alpha_c(512)$ cannot be reliably estimated from only
${\sim}$9 natural-stop samples ($F_L(512) = 0.7\%$);
assuming $\alpha_c$ is approximately model-invariant
(supported by 8B's $\alpha_c(512) = 0.990$, based on 494 samples):
$F_L(b^*) \approx 0.955 / 0.990 = 0.965$---the crossover requires
${\sim}$97\% of chains to complete.
On GSM8K with 27B, whose chains are substantially longer, the crossover is expected well beyond 2048 tokens,
consistent with the observation that thinking has not matched non-thinking at any tested budget.
For 8B, with shorter chains, the crossover occurs at a lower budget but still exceeds 512 tokens.
\end{proof}

\begin{corollary}[Budget multiplier]
\label{cor:multiplier}
The budget multiplier $\gamma \triangleq b^*/b_{\mathrm{sat}}$ quantifies
the token overhead for thinking to become viable.
Since $b_{\mathrm{sat}}$ is determined by answer length (not chain length):
$\gamma = F_L^{-1}(\bar{\alpha}_{\mathrm{nt}}/\alpha_c) \big/ b_{\mathrm{sat}}$.
On MATH-500, nothink accuracy has not fully saturated by $b{=}2048$
(59.8\% at 1024, 64.4\% at 2048), so $b_{\mathrm{sat}} \ge 2048$;
since thinking has not caught non-thinking at $b{=}2048$
(Table~\ref{tab:math500-tax-full}), $b^* > 2048$, confirming
that harder tasks demand larger budgets for thinking to become viable.
For 27B on GSM8K, $b_{\mathrm{sat}} = 512$ (nothink saturates at 95.5\%) and the crossover is not reached at the tested budgets, implying $\gamma \gg 1$.
\end{corollary}

% ---------------------------------------------------------------------------
\subsection{Natural Stop as a Confidence Oracle}
\label{sec:theory:oracle}
% ---------------------------------------------------------------------------

\begin{proposition}[Oracle precision]
\label{prop:oracle}
Under Assumption~\ref{asm:binary}, the natural-stop event
$\mathcal{S}(b) \triangleq \{L(q) \le b\}$
(the chain completes before the budget is exhausted) has positive predictive
value:
\begin{equation}
  \mathrm{PPV}(\mathcal{S}) \;=\;
  \Pr(\mathrm{correct} \mid L \le b)
  \;=\; \alpha_c(b).
  \label{eq:ppv}
\end{equation}
\end{proposition}

\noindent
\textbf{Empirical proxy.}
In practice, we observe the output length $|y|$ rather than the
latent $L(q)$.
Since $|y| = \min(L(q), b)$, we use $|y| < b$ as a proxy for
$\{L \le b\}$.
For discrete token counts, $|y| < b$ corresponds to $L \le b - 1$;
the difference is negligible for our budgets.%
\footnote{Table~\ref{tab:natural-stop-oracle} uses $|y| < 0.95\,b$
as a stricter empirical threshold to exclude borderline cases.
This selects a subset of $\{L \le b\}$; the PPV on this subset
is a separate estimate that may differ from $\alpha_c(b)$.}

\noindent
\textbf{Non-thinking mode.}
Empirically, the natural-stop signal is also highly predictive in
non-thinking mode:
at $b{=}256$ on GSM8K, $\mathrm{PPV}(\mathcal{S}_{\mathrm{nt}}) = 94.4\%$
(Table~\ref{tab:natural-stop-oracle}).
We observe this consistently across budgets and benchmarks, though
unlike the thinking-mode case it does not follow from
Assumption~\ref{asm:binary} alone.

\noindent
By Hoeffding's inequality, with $n{=}475$ natural-stop samples
at the stricter $|y| < 0.95\,b$ threshold (8B, $b{=}512$)
and observed accuracy $\hat{\alpha} = 0.963$ on that subset:%
\footnote{This 96.3\% is the accuracy of the $|y| < 0.95\,b$ subset,
which is slightly lower than the $\alpha_c(512) = 99.0\%$ measured
on \emph{all} samples with $L \le b$ (the broader event).
The difference arises because the stricter threshold selects 475 of 494
natural-stop samples, and the excluded borderline cases happen to be
correct.}
the 95\% lower confidence bound on this proxy PPV
(treating the 475 samples as i.i.d.\ draws from the question
distribution) is
$\hat{\alpha} - \sqrt{\ln(2/0.05)/(2 \cdot 475)} = 0.963 - 0.062 = 0.901$.
This bounds the proxy PPV (on the $|y|<0.95\,b$ subset), not $\alpha_c(b)$
itself, but confirms that the natural-stop signal is reliable even under
finite-sample uncertainty.

\begin{remark}[Information-theoretic interpretation]
Early stopping implies the model's conditional entropy
$H(A \mid y_{1:t^*}) \approx 0$ at $t^* \ll b$---the model is
near-certain of its answer.
Unlike logit-based confidence~\citep{kadavath2022language}, this signal
is purely behavioral: it requires no access to model internals,
no calibration, and no additional compute.
\end{remark}

% ---------------------------------------------------------------------------
\subsection{Inverse Scaling with Model Size}
\label{sec:theory:scaling}
% ---------------------------------------------------------------------------

\begin{proposition}[Inverse scaling of the thinking tax]
\label{prop:inverse-scaling}
Let $L_M$ denote the chain-length distribution for a model of size $M$.
If (i)~non-thinking accuracy $\mathrm{Acc}_{\mathrm{nt}}(b)$ is approximately
size-invariant at moderate budgets,
(ii)~the truncation-dominated regime holds ($\alpha_t \approx 0$),
and (iii)~the completed-chain contribution satisfies
$F_{L_{M_2}}(b) \cdot \alpha_c(M_2, b) \le F_{L_{M_1}}(b) \cdot \alpha_c(M_1, b)$
for $M_2 > M_1$,
then for any fixed budget $b$:
\begin{equation}
  \mathrm{Tax}(M_2, b) \;\ge\; \mathrm{Tax}(M_1, b).
  \label{eq:inverse-scaling}
\end{equation}
Condition~(iii) is implied by stochastic dominance of $L_{M_2}$ over $L_{M_1}$
when $\alpha_c$ is approximately model-invariant, and is verified
empirically: at $b{=}512$,
$F_{L_{\text{8B}}} \alpha_c = 0.370$ vs.\
$F_{L_{\text{27B}}} \alpha_c = 0.007$.
\end{proposition}

\begin{proof}
From Proposition~\ref{thm:thinking-tax} with $\alpha_t \approx 0$ (condition~(ii)):
$\mathrm{Tax}(M, b) \approx \mathrm{Acc}_{\mathrm{nt}}(b) -
F_{L_M}(b) \cdot \alpha_c(M, b)$.
By condition~(i), the first term is approximately constant across $M$.
By condition~(iii),
$F_{L_{M_2}}(b) \cdot \alpha_c(M_2, b)
\le F_{L_{M_1}}(b) \cdot \alpha_c(M_1, b)$,
so the second term is smaller for the larger model, increasing the tax.
\end{proof}

\noindent
\textbf{Empirical verification.}
At $b{=}512$: 8B natural-stop rate = 37.4\%, 9B $\approx$ 0.5\%, 27B = 0.7\%.
Meanwhile, non-thinking accuracy at budget 512 is uniformly high: 8B = 93.1\%, 9B = 93.2\%, 27B = 95.5\%.
The thinking tax for 9B is $93.2\% - 15.5\% = 77.7$\,pp and for 27B is $95.5\% - 18.3\% = 77.2$\,pp, vs.\
8B's $93.1\% - 56.9\% = 36.2$\,pp---a ${\sim}2.1\times$ amplification from 8B to 9B,
as predicted by stochastic dominance of chain-length distributions.

\begin{corollary}[Crossover budget grows with model size]
\label{cor:crossover-scaling}
Under the heuristic (Eq.~\ref{eq:crossover}), if
$\bar{\alpha}_{\mathrm{nt}}$ and $\alpha_c$ are approximately
model-invariant and $L_{M_2}$ stochastically dominates $L_{M_1}$
(i.e., $F_{L_{M_2}}(b) \le F_{L_{M_1}}(b)$ for all $b$),
then
$b^*(M) \approx F_{L_M}^{-1}(\bar{\alpha}_{\mathrm{nt}}/\alpha_c)$
shifts right with $M$:
$M_2 > M_1 \implies b^*(M_2) \ge b^*(M_1)$.
Larger models require proportionally larger budgets for thinking to
become cost-effective.
\end{corollary}

% ---------------------------------------------------------------------------
\subsection{From Truncation Waste to Split-Budget Generation}
\label{sec:theory:bridge}
% ---------------------------------------------------------------------------

The decomposition framework reveals a \emph{major structural cause} of
the thinking tax: the coupling constraint $|Z| + |A| \le b$ forces
reasoning and answering to compete for the same token budget.%
\footnote{When $\alpha_c \approx \bar{\alpha}_{\mathrm{nt}}$ (completed
thinking is as accurate as non-thinking), truncation waste accounts for
essentially the entire tax.
If $\alpha_c < \bar{\alpha}_{\mathrm{nt}}$ (thinking is intrinsically
less accurate even when completed), a residual tax persists beyond what
decoupling can eliminate.
On GSM8K at $b{=}512$, $\alpha_c = 99.0\% > \bar{\alpha}_{\mathrm{nt}}
= 93.1\%$, so truncation is the dominant mechanism in our setting.}
This suggests a direct remedy: \emph{decouple} the two stages by
allocating separate budgets.

\begin{remark}[From diagnosis to intervention]
\label{rem:split-bridge}
The accuracy decomposition (Eq.~\ref{eq:acc-decomp}) identifies two
levers for improving thinking-mode performance at budget $b$:
(i)~increase $F_L(b)$ (allow more chains to complete), or
(ii)~increase $\alpha_t(b)$ (extract more value from truncated chains).
Lever~(i) requires either larger budgets or shorter chains (model retraining).
Lever~(ii) can be achieved at inference time: by feeding a truncated
reasoning trace to a separate answer-generation pass, we replace
$\alpha_t(b)$ with $\alpha_{\mathrm{extract}}(b) \gg \alpha_t(b)$,
recovering value that would otherwise be lost to truncation waste.
\end{remark}

\noindent
This observation motivates \emph{split-budget generation}
(\S\ref{sec:method:split}): rather than attempting to prevent
truncation, we \emph{accept} it and recover its value through a
decoupled answer pass.
The truncation-waste framework provides the theoretical foundation
(identifying the coupling constraint as the dominant mechanism behind the
tax when $\alpha_c \ge \bar{\alpha}_{\mathrm{nt}}$), while the split-budget framework provides the operational remedy
(eliminating the constraint by separating reasoning and answering
into independent passes).

% ---------------------------------------------------------------------------
\subsection{The Coupling Tax as a Rate Constraint}
\label{sec:theory:coupling}
% ---------------------------------------------------------------------------

We now formalize the coupling constraint, yielding a tighter
characterization of \emph{why} split-budget generation helps and
\emph{how much} it can recover.
The following results are consequences of the decomposition framework
above; their value lies in making \emph{quantitative predictions} about
the gains from decoupling, which we verify empirically.

\begin{definition}[Coupled vs.\ split generation]
\label{def:coupled-split}
In \emph{coupled} generation, reasoning tokens~$Z$ and answer tokens~$A$
share a single budget: $|Z| + |A| \le b$.
In \emph{split-budget} generation, reasoning receives budget~$b_r$ and
answering receives an independent budget~$b_a$, with no constraint
coupling the two.
\end{definition}

\noindent
The coupling constraint $|Z|+|A| \le b$ acts as a \emph{rate limit}
on a joint channel: reasoning and answering compete for shared bandwidth.
When the reasoning trace is long ($|Z| \gg b - b_{\min}$, where $b_{\min}$
is the minimum tokens needed for a well-formed answer), the answer is
forcibly truncated or absent---not because the model lacks the answer, but
because the channel has no remaining capacity.
Split-budget generation removes this contention by allocating
independent channels.

\begin{proposition}[Recoverable coupling tax]
\label{prop:recoverable-tax}
Let $\alpha_{\mathrm{extract}}(b_r, b_a)$ denote the accuracy of a
separate answer-extraction pass that receives the first $b_r$ tokens
of the reasoning trace and generates up to $b_a$ answer tokens.
Assume that chains which complete within $b_r$ retain their coupled
accuracy $\alpha_c(b_r)$---i.e., completed chains are used directly
without re-extraction.%
\footnote{In practice, this is implemented by checking for a
natural-stop signal: if the thinking chain terminates before $b_r$,
its answer is used directly; extraction applies only to truncated chains.
Without this bypass, completed chains could be harmed by the extraction
pass, invalidating the decomposition.}
Compared to coupled generation at the \emph{same reasoning budget}~$b_r$,
the accuracy gain from split-budget generation is:
\begin{equation}
  \Delta_{\mathrm{split}}(b_r, b_a)
  \;=\;
  \bigl(1 - F_L(b_r)\bigr) \cdot
  \bigl(\alpha_{\mathrm{extract}}(b_r, b_a) - \alpha_t(b_r)\bigr).
  \label{eq:recoverable-tax}
\end{equation}
This quantity is non-negative whenever
$\alpha_{\mathrm{extract}} \ge \alpha_t$---i.e., whenever the
decoupled answer pass extracts at least as much value as the truncated
coupled output---and grows with the truncation rate $(1 - F_L(b_r))$.

When comparing to a coupled baseline at a \emph{larger} total budget
$b = b_r + b_a$, an additional correction applies.
Writing $\alpha_c(b_r)$ and $\alpha_c(b)$ for the completed-chain
accuracy at the respective budgets:
\begin{equation}
  \Delta_{\mathrm{split}}^{\mathrm{total}}
  \;=\;
  \bigl(1 - F_L(b_r)\bigr)
    \bigl(\alpha_{\mathrm{extract}} - \alpha_t(b)\bigr)
  \;-\;
  \bigl(F_L(b) - F_L(b_r)\bigr)
    \bigl(\alpha_c(b) - \alpha_t(b)\bigr)
  \;+\;
  F_L(b_r)\bigl(\alpha_c(b_r) - \alpha_c(b)\bigr).
  \label{eq:recoverable-tax-total}
\end{equation}
The second term is non-positive (more chains complete at higher budget),
and the third term captures the difference in completed-chain accuracy
at the two budgets (typically small: $\alpha_c(b_r) \approx \alpha_c(b)$
for moderate $b_a$).
Since $\Delta^{\mathrm{total}}_{\mathrm{split}}$ is the sum of all
three terms, split-budget generation is advantageous at matched total
budget when the positive first term (truncation recovery) outweighs the
sum of the non-positive second term and the (typically small) third term.
In our experiments, $b_a = 256$ and
$\alpha_c(b_r) \approx \alpha_c(b_r + 256)$, so the third term
is negligible and the sufficient condition reduces to
the first term exceeding the magnitude of the second.
\end{proposition}

\begin{proof}
Split-budget accuracy decomposes as:
$\mathrm{Acc}_{\mathrm{split}}(b_r, b_a)
= F_L(b_r) \cdot \alpha_c(b_r)
+ (1 - F_L(b_r)) \cdot \alpha_{\mathrm{extract}}(b_r, b_a)$.
For Eq.~\eqref{eq:recoverable-tax}, subtract the coupled accuracy
(Eq.~\ref{eq:acc-decomp}) at budget~$b_r$:
$\Delta_{\mathrm{split}}
= (1 - F_L(b_r))(\alpha_{\mathrm{extract}} - \alpha_t(b_r))$,
which is non-negative whenever $\alpha_{\mathrm{extract}} \ge \alpha_t(b_r)$.
For Eq.~\eqref{eq:recoverable-tax-total}, subtract the coupled
accuracy at $b = b_r + b_a$:
$\Delta^{\mathrm{total}}
= \bigl[F_L(b_r) \alpha_c(b_r) + (1{-}F_L(b_r))\alpha_{\mathrm{extract}}\bigr]
  - \bigl[F_L(b) \alpha_c(b) + (1{-}F_L(b))\alpha_t(b)\bigr]$.
Rearranging yields the three terms in Eq.~\eqref{eq:recoverable-tax-total}.
\end{proof}

\noindent
\textbf{Empirical verification.}
Eq.~\eqref{eq:recoverable-tax} compares split-budget to coupled generation
at the same reasoning budget.
On MATH-500, IRIS uses $b_r \in \{2048, 4096\}$ with $b_a = 256$.
At $b_r{=}2048$: $1 - F_L(2048) = 0.912$ (91.2\% truncated), and IRIS
achieves 67.2\% vs.\ TOWN's 55.0\% ($\Delta = +12.2$\,pp, $p < 10^{-6}$).
At $b_r{=}4096$: $1 - F_L(4096) = 0.553$, and the gap narrows to
$+2.2$\,pp ($p = 0.28$), consistent with the prediction that lower
truncation rate $\implies$ smaller $\Delta_{\mathrm{split}}$.
The ratio of truncation rates ($0.912/0.553 = 1.65$) directionally matches
the ratio of observed IRIS--TOWN gaps, confirming that the coupling tax
drives IRIS's advantage.
For the matched-total-budget comparison (Eq.~\ref{eq:recoverable-tax-total}),
the correction terms are small
when $b_a \ll b_r$ (as in our experiments with $b_a = 256$), because
$F_L(b_r + 256) \approx F_L(b_r)$ for the heavy-tailed chain-length
distributions observed in practice.

\begin{proposition}[Optimal budget allocation]
\label{prop:optimal-split}
Given a total token budget $B$ and the choice to allocate
$b_r$ tokens to reasoning and $b_a = B - b_r$ to answering,
the optimal reasoning budget is:
\begin{equation}
  b_r^* \;=\; \arg\max_{b_r \in [0, B]} \;
  F_L(b_r) \cdot \alpha_c(b_r)
  + \bigl(1 - F_L(b_r)\bigr) \cdot
    \alpha_{\mathrm{extract}}(b_r, B - b_r).
  \label{eq:optimal-split}
\end{equation}
The maximizer $b_r^*$ balances two competing effects:
increasing $b_r$ raises the completion rate $F_L(b_r)$ but reduces
the answer budget $B - b_r$.
In our experiments, $b_a = 256$ tokens suffices for answer extraction
(MATH-500 answers average $<$50 tokens), so the optimum allocates
most of the budget to reasoning ($b_r^* \approx B - 256$).
\end{proposition}

\noindent
The intuition is a bias--variance tradeoff: allocating more tokens to
reasoning increases the completion rate~$F_L(b_r)$ (reducing the
``truncation rate'' and its associated accuracy penalty), but leaves fewer
tokens for the answer pass when truncation does occur.
The optimum balances these competing demands.
In practice, our IRIS experiments use fixed $b_a = 256$ (sufficient for
answer extraction on MATH-500), allocating the remaining budget to reasoning.

\noindent
\textbf{Cross-architecture generality.}
The coupling constraint $|Z|+|A| \le b$ is not model-specific: it
applies to any autoregressive generator that produces reasoning and
answers in a single pass.
On DeepSeek-R1-Distill-Llama-8B (GSM8K, $n{=}1{,}319$), the truncation
mechanism replicates: at $b{=}256$, 95.8\% of chains are truncated (accuracy 7.5\%);
at $b{=}1024$, 99.6\% complete naturally ($\alpha_c = 0.647$).
On MATH-500 ($n{=}500$), natural-stop rates rise from 16.8\% at $b{=}512$
to 76.0\% at $b{=}4096$, mirroring the Qwen pattern.
However, when comparing thinking vs.\ prompt-based nothink on DeepSeek,
the \emph{tax magnitude} is near-zero (average +0.3\,pp across MATH-500 budgets;
Appendix~\ref{app:deepseek}), because the prompt-based nothink does not
produce meaningfully different generation behavior.
This highlights that the coupling tax requires two conditions:
(1)~truncation waste from shared budgets (universal, verified on both
architectures) and (2)~a genuinely distinct non-thinking mode
(model-dependent---native in Qwen, absent in DeepSeek).
The truncation mechanism is a \emph{structural property of the generation format};
the tax magnitude depends on the availability of an efficient alternative mode.
