% thinking_tax_final.tex — §3 The Thinking Tax: Empirical Analysis
% Presents the core empirical findings

\subsection{Setup}
\label{sec:tax-setup}

We evaluate the \thinkingtax{} across two dimensions: (i)~\emph{matched budgets}, where thinking and non-thinking modes operate under identical token limits, and (ii)~\emph{actual token efficiency}, measuring accuracy relative to tokens consumed.

\paragraph{Models and Benchmarks.}
We use \textbf{Qwen3-8B} and \textbf{Qwen3.5-27B}~\citep{qwen2025qwen3} on GSM8K~\citep{cobbe2021gsm8k} ($n$=1,319 test problems).
Both models support a native \texttt{enable\_thinking} flag that toggles generation of \texttt{<think>...</think>} reasoning blocks.
We additionally validate on MATH-500 with DeepSeek-R1-Distill-Llama-8B~\citep{deepseek2025r1}.

\paragraph{Non-thinking mode.}
With \texttt{enable\_thinking=False}, the model generates a direct answer without any reasoning prefix.
We sweep budgets $B \in \{32, 64, 128, 256, 512\}$.

\paragraph{Thinking mode.}
With \texttt{enable\_thinking=True}, the model first generates a reasoning trace, then a final answer.
We test $B \in \{128, 256, 512\}$.
When a response exceeds the budget, we apply projection---a simple extraction heuristic that searches the truncated output for a numerical answer.

\subsection{Finding 1: Non-Thinking Dominates at Every Budget}
\label{sec:finding1}

\input{../results/paper_figures/table_thinking_tax_main}

Table~\ref{tab:thinking-tax} presents the complete results on the full test set ($n{=}1{,}319$).
At \textbf{every tested budget}, non-thinking mode achieves dramatically higher accuracy:

\begin{itemize}[nosep,leftmargin=*]
    \item \textbf{Budget 128}: nothink 50.8\% vs.\ think 3.0\% (+47.8\,pp)
    \item \textbf{Budget 256}: nothink 87.5\% vs.\ think 18.0\% (+69.5\,pp)
\end{itemize}

The headline comparison: \texttt{nothink@256} achieves \textbf{87.5\%} accuracy with only 146 avg tokens, compared to \texttt{think@512}'s 65.2\% at 477 avg tokens---non-thinking is \textbf{22.3\,pp more accurate} while consuming \textbf{3.3$\times$ fewer tokens}.
For the 27B model, the tax is even more dramatic: \texttt{nothink@512} = 95.5\% vs.\ \texttt{think@512} = 18.3\%, a \textbf{77.2\,pp} gap.

\subsection{Finding 2: Early Stopping as Confidence Oracle}
\label{sec:finding2}

\input{../results/paper_figures/table_natural_stop_oracle}

A key structural insight: \textbf{whether a model stops before exhausting its budget is a highly predictive signal of correctness} (Table~\ref{tab:natural-stop-oracle}).

For non-thinking mode at budget 256, 88.8\% of samples (1{,}171/1{,}319) stop early (using ${\sim}$133 avg tokens), and these early-stopped samples achieve 94.4\% accuracy.
For thinking mode at budget 512 (8B), samples that stop early achieve a remarkable \textbf{93.8\%} accuracy (among 37.4\% that complete naturally).
This natural-stop signal is \emph{free}---it requires no auxiliary model, no additional computation, and no training.

\subsection{Finding 3: Token Waste Scales with Budget}
\label{sec:finding3}

\input{../results/paper_figures/table_token_utilization}

Token utilization---the fraction of the budget actually consumed---decreases systematically with budget (Table~\ref{tab:token-utilization}).
For non-thinking mode on the full test set ($n{=}1{,}319$):
\begin{itemize}[nosep,leftmargin=*]
    \item Budget 128: 88.4\% utilization (113/128 tokens)
    \item Budget 256: 57.0\% utilization (146/256 tokens)
\end{itemize}
This drop reveals that non-thinking answers are \emph{inherently compact}: the model needs ${\sim}$146 tokens regardless of the budget allocated.
The implication is that budgets beyond ${\sim}$256 provide diminishing returns for non-thinking mode on GSM8K.
