\documentclass{article}

\usepackage[final]{neurips_2025}
\makeatletter\renewcommand{\@neuripsyear}{2026}\makeatother

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{colortbl}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{tablefootnote}
\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,positioning,fit,calc}

\definecolor{improve}{RGB}{0,128,0}
\definecolor{decline}{RGB}{200,0,0}
\definecolor{easycolor}{RGB}{76,175,80}
\definecolor{medcolor}{RGB}{255,152,0}
\definecolor{hardcolor}{RGB}{244,67,54}
\newcommand{\up}[1]{\textcolor{improve}{\textbf{+#1}}}
\newcommand{\down}[1]{\textcolor{decline}{\textbf{#1}}}
\newcommand{\ci}[2]{\scriptsize[#1,\,#2]}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\method}{\textsc{ReasonSpec}}
\newcommand{\fullmethod}{Reasoning Speculation}

\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\theoremstyle{remark}
\newtheorem{remark}{Remark}

% NeurIPS checklist
\providecommand{\answerYes}[1]{\textbf{Yes.} #1}
\providecommand{\answerNo}[1]{\textbf{No.} #1}
\providecommand{\answerNA}[1]{\textbf{N/A.} #1}

\title{\fullmethod{}: Parallel Exploration with Cross-Path\\Consensus for Adaptive Test-Time Compute}

\author{
  Anonymous Author(s)
}

\begin{document}

\maketitle

\begin{abstract}
% PLACEHOLDER — will be finalized after experiments complete
Scaling test-time compute via chain-of-thought reasoning improves accuracy but wastes tokens on easy problems and underallocates on hard ones.
Existing adaptive methods rely on single-path difficulty signals (answer presence, uncertainty, learned features) which we show are fundamentally unreliable: all such black-box controllers fail to outperform fixed budgets.
We introduce \method{}, a training-free framework that restructures the computation of reasoning itself.
Instead of allocating a single token budget and hoping it matches problem difficulty, \method{} generates $K$ short parallel exploration paths, computes a \emph{cross-path consensus signal}, and adaptively routes each problem:
high consensus $\rightarrow$ accept the majority answer (easy); partial consensus $\rightarrow$ extend the most promising path (medium); low consensus $\rightarrow$ full deliberation with all paths as context (hard).
The key insight is that multi-path agreement is a far more reliable difficulty signal than any feature extractable from a single reasoning trace (Spearman $\rho = -0.51$ vs.\ difficulty, $p < 10^{-88}$).
On GSM8K with Qwen3-8B, \method{} achieves \textbf{XX.X\%} accuracy using only \textbf{XXX} average tokens, compared to \textbf{XX.X\%} for a fixed budget using the same total compute, representing a \textbf{+X.X\,pp} improvement at \textbf{XX\%} token savings.
\method{} is model-agnostic, requires no training, and works at the API level with any reasoning LLM.
\end{abstract}

% ============================================================================
\section{Introduction}
\label{sec:intro}
% ============================================================================

\input{sections/intro_reasonspec}

% ============================================================================
\section{Related Work}
\label{sec:related}
% ============================================================================

\input{sections/related_reasonspec}

% ============================================================================
\section{Method: \fullmethod{}}
\label{sec:method}
% ============================================================================

\input{sections/method_reasonspec}

% ============================================================================
\section{Theoretical Analysis}
\label{sec:theory}
% ============================================================================

\input{sections/theory_reasonspec}

% ============================================================================
\section{Experiments}
\label{sec:experiments}
% ============================================================================

\input{sections/experiments_reasonspec}

% ============================================================================
\section{Analysis}
\label{sec:analysis}
% ============================================================================

\input{sections/analysis_reasonspec}

% ============================================================================
\section{Conclusion}
\label{sec:conclusion}
% ============================================================================

\input{sections/conclusion_reasonspec}


\bibliographystyle{plain}
\bibliography{references}

% ============================================================================
\appendix
\input{sections/appendix_reasonspec}

\end{document}
