\documentclass{article}

\usepackage[final]{neurips_2025}
\makeatletter\renewcommand{\@neuripsyear}{2026}\makeatother

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{colortbl}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{multirow}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{tablefootnote}
\usepackage{enumitem}

\definecolor{improve}{RGB}{0,128,0}
\definecolor{decline}{RGB}{200,0,0}
\newcommand{\up}[1]{\textcolor{improve}{\textbf{+#1}}}
\newcommand{\down}[1]{\textcolor{decline}{\textbf{#1}}}
\newcommand{\ci}[2]{\scriptsize[#1,\,#2]}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\adathink}{\textsc{AdaThink}}

% NeurIPS checklist answer macros (only define if not already in .sty)
\providecommand{\answerYes}[1]{\textbf{Yes.} #1}
\providecommand{\answerNo}[1]{\textbf{No.} #1}
\providecommand{\answerNA}[1]{\textbf{N/A.} #1}

\title{\adathink{}: Adaptive Budget Allocation for\\Test-Time Compute in Large Language Models}

\author{
  Anonymous Author(s)
}

\begin{document}

\maketitle

\begin{abstract}
Scaling test-time compute via extended chain-of-thought reasoning improves LLM accuracy on hard problems, but we show that this improvement is far from uniform: fixed budgets \emph{systematically overthink easy questions and undercompute hard ones}, creating structured heterogeneity that simple post-hoc controllers can exploit.
We introduce \adathink{}, a training-free, inference-only framework that allocates per-question reasoning tokens based on difficulty signals extracted from a single low-budget probe pass---requiring no model modification or fine-tuning.
Our primary controller is a lookup table over three binary features (answer presence, token utilization, answer consistency) optimized via leave-one-subset-out cross-validation.
Despite its simplicity, this 3-bit template closes $76\%$ of the gap to an oracle budget selector on GSM8K-27B.
Evaluated across three benchmarks (GSM8K, MATH500, BIG-Bench Hard) and two Qwen model scales (8B and 27B), the template controller achieves \textbf{+11.3--28.9}\,pp accuracy gains over matched fixed-budget baselines, with all CIs excluding zero.
The two-pass overhead is offset by accuracy improvements, yielding positive utility gains ($+7.4$--$23.1$\,pp) across all settings.
Feature-stratified analysis reveals the mechanism: $\sim$13\% of questions are ``easy'' (60\% overthinking rate at the max budget), while $\sim$60\% are ``hard'' and benefit from budget escalation.
Cross-benchmark transfer shows the difficulty pattern generalizes: templates learned on MATH500 transfer perfectly to BBH.
\end{abstract}

\input{sections/introduction}
\input{sections/related}
\input{sections/method}
\input{sections/experiments}
\input{sections/conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\input{sections/checklist}

\appendix
\input{sections/appendix}

\end{document}
